Light inextensible strings under tension in the Schwarzschild geometry

### 2019-08-28 string_physics_teacher.tex
work on the AIP style manual

Light inextensible string under tension is a stalwart feature of
elementary physics.  Here I show how considering such string in the
vicinity of a black hole, together with computer algebra systems, can
generate insight into the Schwarzschild geometry that is accessible to
undergraduates.  Light taut strings minimize their proper length,
given by integrating the spatial component of the Schwarzschild metric
along the string.  The path itself is given by straightforward
numerical solution to the Euler-Lagrange equations.  If the string is
entirely outside the event horizon, its closest approach to the
singularity is tangential.  At this point the string is {\em visibly}
curved, surely a memorable and informative insight.  The geometry of
the Schwarzschild metric induces some peculiar nonlocal behaviour: if
the distance of closest approach is less than about $1.0761\cdot 2M$,
the string self-intersects, even though it is everywhere under
tension.  Light taut strings furnish a third interpretation of the
concept ``straight line'', the other two being null geodesics and
free-fall world lines.

@@

\noindent where dashes denote differentiation with respect to~$\phi$.  The
Euler-Lagrange equations for this system,
$\frac{d}{d\phi}\frac{\partial F}{\partial r'}-\frac{\partial
  F}{\partial r}=0$ give, after simplification\footnote{The Mathematica
  idiom would be:

\noindent
{\tt
Needs["VariationalMethods`"]\\
EulerEquations[Sqrt[y'[x]\^{}2/(1-1/y[x]) + y[x]\^{}2],y[x],x]}}, an expression for the second derivative of~$r$:

@@

There does not appear to be a simple analytical solution to
equation~\ref{rdashdash} but nevertheless we may make several
observations.  Equation~\ref{rdashdash} is a second-order nonlinear
ordinary differential equation and thus has two constants of
integration.  There are two qualitatively different types of
solutions: those with a tangential point ($r'=0,r>2M$), and those
without.  Consider first solutions with such a point: there exists a
pair~$\left(r,\phi\right), r>2M$ at which~$r'(\phi)=0$.  This point
must be unique, as~$r''(\phi)>0$, and be the closest approach of the
string to the event horizon.

@@

Here I solve equation~\ref{rdashdash} numerically, using R programming
language~\cite{rcore2019} and the \verb+deSolve+ package for ordinary
differential equations~\cite{soetart2010}.  Non-radial solutions fall
into two distinct classes: those with $r>2M$ at all points, and those
which attain $r=2M$, in which case the string is tangential to the
event horizon.

We first consider strings for which $r>2M$ everywhere (that is,
completely exterior strings).
Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rdashdash}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings
(Figure~\ref{light_closest_approach} shows equivalent null geodesics
for comparison).  A bisection technique reveals that if the closest
approach is greater than a critical value of about~$\simeq 1.0761\cdot
2M$, the string does not self-intersect; if the closest approach is
less than this value, then the string crosses itself at~$\phi=\pi/2$.

There are other ways to parametrize completely exterior strings;
Figure \ref{strings_r_equals_2}, for example, shows strings passing
through~$(4M,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
comparison.

The cases where the string attains $r=2M$ are tangent to the event
horizon (or, exceptionally, purely radial).
Figure~\ref{fixed_EH_intersection} shows strings tangential to the
event horizon at~$\phi=0$ and Figure~\ref{one_free_end_r_equals_2}
shows tangential strings passing through $(4M,0)$.

@@

\section*{Supplementary Material}

Calculations are performed using the R programming language~\cite{rcore2018};
source code for the figures is available at
\\
\\
{\tt https://github.com/RobinHankin/string.git}

@@

\begin{figure}[p] % Created by "closest_approach.R"
\centering
\includegraphics[width=\linewidth]{closest_approach_nonselfintersecting.pdf}
\caption{Non-self intersecting strings arranged by distance~$d$ of
  closest approach to the event horizon, occurring tangentially
  at~$\phi=0$}
\label{closest_approach_non_self_intersecting}
\end{figure}

\begin{figure}[p] % created by "closest_approach2.R"
\centering
\includegraphics[width=\linewidth]{closest_approach_selfintersecting.pdf}
\caption{Self-intersecting strings shown for $\phi\geqslant 0$.  Closest
  approach to the event horizon occurs at $\phi=0$ (3 o'clock) and
  $r/2M\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$
  but for clarity only $\phi\geqslant 0$ is shown}
\label{closest_approach_self_intersecting}
\end{figure}

\begin{figure}[p] % Created by "light_closest_approach.R"
\centering
\includegraphics[width=\linewidth]{light_closest_approach.pdf}

\caption{Null geodesics in the Schwarzschild geometry, tangential
  at~$\phi=0$.  Note the differences between these curves and the taut
  strings shown elsewhere: unlike taut strings, null geodesics have
  positive curvature everywhere, may cross the event horizon inwards,
  and are never tangential to the event horizon}
\label{light_closest_approach}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=\linewidth]{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
\caption{Strings passing through $(4M,0)$}
\label{strings_r_equals_2}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=\linewidth]{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
\caption{Null geodesics passing through $(4M,0)$, shown for
  $0\leqslant\phi\leqslant 2\pi$}
\label{light_r_equals_2}
\end{figure}

@@

 \section*{Conflict of interest}
 The author declares that he has no conflict of interest.

\subsection*{Supplementary material}

R code for all statistical analysis given here is available at

{\tt https://github.com/RobinHankin/hyper2}.

\section*{References}

\bibliography{stringrefs.bib}

>>>>

Light inextensible string under tension is a stalwart feature of
elementary physics.  Here I show how considering such string in the
vicinity of a black hole, together with computer algebra systems, can
generate insight into the Schwarzschild geometry that is suitable for
an undergraduate homework problem.  Light taut strings minimize their
proper length, given by integrating the spatial component of the
Schwarzschild metric along the string.  The path itself is given by
straightforward numerical solution to the Euler-Lagrange equations.
If the string is entirely outside the event horizon, its closest
approach to the singularity is tangential.  At this point the string
is {\em visibly} curved, surely a memorable and informative insight.
The geometry of the Schwarzschild metric induces some peculiar
nonlocal behaviour: if the distance of closest approach is less than
about $1.0761\cdot 2M$, the string self-intersects, even though it is
everywhere under tension.  Light taut strings furnish a third
interpretation of the concept ``straight line'', the other two being
null geodesics and free-fall world lines.

@@

\noindent where dashes denote differentiation with respect to~$\phi$.
The Euler-Lagrange equations for this system,
$\frac{d}{d\phi}\frac{\partial F}{\partial r'}-\frac{\partial
  F}{\partial r}=0$ give an expression for the second derivative
of~$r$ (Mathematica idiom would be {\tt EulerEquations[Sqrt[y[x]\^{}2
      + y'[x]\^{}2/(1-1/y[x])],y[x],x]}):

@@

There does not appear to be a simple analytical solution to
equation~\ref{rdashdash} but it is a straightforward second order ODE,
amenable to numerical solution.  There are two qualitatively different
types of solutions: those with a tangential point ($r'=0,r>2M$), and
those without.  Consider first solutions with such a point: there
exists a pair~$\left(r,\phi\right), r>2M$ at which~$r'(\phi)=0$.  This
point must be unique, as~$r''(\phi)>0$, and be the closest approach of
the string to the event horizon.

@@

Here I solve equation~\ref{rdashdash} numerically, using the R
programming language~\cite{rcore2019} and the \verb+deSolve+ package
for ordinary differential equations~\cite{soetart2010}.  First we
consider strings for which $r>2M$ everywhere (that is, completely
exterior strings).
Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rdashdash}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings
(Figure~\ref{light_closest_approach} shows equivalent null geodesics
for comparison).  A bisection technique reveals that if the closest
approach is greater than a critical value of about~$\simeq 1.0761\cdot
2M$, the string does not self-intersect; if the closest approach is
less than this value, then the string crosses itself at~$\phi=\pi/2$.

There are other ways to parametrize completely exterior strings;
Figure \ref{strings_r_equals_2}, for example, shows strings passing
through~$(4M,0)$.  Figure~\ref{light_r_equals_2} shows light paths for
comparison.  The cases where the string attains $r=2M$ are tangent to
the event horizon (or, exceptionally, purely radial).
Figure~\ref{fixed_EH_intersection} shows strings tangential to the
event horizon at~$\phi=0$ and Figure~\ref{one_free_end_r_equals_2}
shows tangential strings passing through $(4M,0)$.

@@

\section*{References}
\bibliography{stringrefs.bib}

\clearpage

@@

\begin{figure}[p] % Created by "closest_approach.R"
\centering
\includegraphics[width=\linewidth]{closest_approach_nonselfintersecting.pdf}
\caption{Light inextensible strings under tension close to a black
  hole.  Here we see non-self intersecting strings arranged by
  increasing distance of closest approach to the event horizon,
  occurring tangentially at~$\phi=0$}
\label{closest_approach_non_self_intersecting}
\end{figure}

\begin{figure}[p] % created by "closest_approach2.R"
\centering
\includegraphics[width=\linewidth]{closest_approach_selfintersecting.pdf}
\caption{Light inextensible strings under tension close to a black
  hole.  Here self-intersecting strings are shown for $\phi\geqslant
  0$.  Closest approach to the event horizon occurs at $\phi=0$ (3
  o'clock) and $r/2M\in\left(1,1.0761\right)$; strings are symmetrical
  about $\phi=0$ but for clarity only $\phi\geqslant 0$ is shown}
\label{closest_approach_self_intersecting}
\end{figure}

\begin{figure}[p] % Created by "light_closest_approach.R"
\centering
\includegraphics[width=\linewidth]{light_closest_approach.pdf}
\caption{Null geodesics in the Schwarzschild geometry, tangential
  at~$\phi=0$.  Note the differences between these curves and the taut
  strings shown elsewhere: unlike taut strings, null geodesices may
  cross the event horizon inwards, and are never tangential to the
  event horizon}
\label{light_closest_approach}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=\linewidth]{angle_at_r_equals_2.pdf} % created by "angle_at_r_equals_2.R"
\caption{Light inextensible strings under tension close to a black
  hole.  Here strings passing through $(4M,0)$ at different angles are
  shown}
\label{strings_r_equals_2}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=\linewidth]{light_start_at_r_equals_2.pdf} % created by "angle_start_at_r_equals_2.R"
\caption{Null geodesics passing through $(4M,0)$ at different angles,
  shown for $0\leqslant\phi\leqslant 2\pi$}
\label{light_r_equals_2}
\end{figure}

@@
